---
title: "MFAssignR Molecular Formula Assignment"
subtitle: "Emily's Metabolomics Data"
date: "`r Sys.Date()`"
output: html_document
---

**Citation (REQUIRED):**
Schum SK, Brown LE, Mazzoleni LR. MFAssignR: Molecular formula assignment
software for ultrahigh resolution mass spectrometry analysis of environmental
complex mixtures. *Environmental Research*, 191, 2020.

**Workflow:** Lucia Liu's MFAssignR guide (05/01/25).
**Parameters:** MFAssignR package defaults (from GitHub `skschum/MFAssignR`).
Code from Dr. Amna Ijaz and Dr. Gregory Vandergrift.

---

# Section 0: Installation

Run once. Not needed every time.

```{r install, eval=FALSE}
library(devtools)
devtools::install_github("skschum/MFAssignR")
```

```{r setup, message=FALSE, warning=FALSE}
library(MFAssignR)
library(readxl)
library(dplyr)
library(ggplot2)

knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.height = 5, fig.width = 7,
                      paged.print = FALSE, results = TRUE)
```

---

# Section 1: File Setting and Noise Prescreening

(Lucia slides 8-9)

```{r load-data}
## Load Emily's data from Excel
## The original guide reads CSV; we read Excel and convert to the same format
df_raw <- read_excel("Emily_Data_Pruned_Labeled.xlsx")
cat("Loaded", nrow(df_raw), "rows\n")

## MFAssignR expects: column 1 = mass, column 2 = abundance
Data_all <- df_raw %>%
  select(`m/z`, `Maximum Abundance`, `Retention time (min)`, Charge) %>%
  rename(mass = `m/z`, Abundance = `Maximum Abundance`, RT = `Retention time (min)`) %>%
  filter(!is.na(mass), !is.na(Abundance), Abundance > 0) %>%
  arrange(mass)

cat("Total input peaks:", nrow(Data_all), "\n")

## Split by charge state for two-pass workflow
## 24% of Emily's peaks are z=2 — running them with Zx=1 gives wrong formulas
Data_z1 <- Data_all %>% filter(Charge == 1) %>% select(-Charge)
Data_z2 <- Data_all %>% filter(Charge == 2) %>% select(-Charge)
cat("z=1 peaks:", nrow(Data_z1), "\n")
cat("z=2 peaks:", nrow(Data_z2), "\n")

## Sections 2-9 below run on z=1 peaks first
## Section 10 at the end runs the same pipeline for z=2 peaks (with Zx=2)
Data <- Data_z1
Data_4 <- Data
```

**NOTE from Lucia (slide 9):** The script may contain a chunk that removes noise
based on standard deviation. This is NOT recommended. We skip it and use
KMDNoise instead (the proper method).

---

# Section 2: KMD Noise Assessment

(Lucia slides 10-14)

## 2a: KMD Noise - Describe and Plot

(Lucia slide 12, step 1)

Set the slice boundaries. These define a region in the KMD plot where only
noise exists (the "sea"), avoiding the analyte "islands".

**USER INPUT:** Look at the plot. If the red slice line cuts through analyte
islands, adjust `upper.y`, `lower.y`, `upper.x`, `lower.x`.

```{r kmd-noise}
## USER INPUT: Describe and plot KMD Noise
## Adjust slice boundaries if needed to avoid analyte islands
Noise <- KMDNoise(Data_4)
plot <- Noise[["KMD"]]
plot
KMDN <- Noise[["Noise"]]
cat("KMD Noise level (KMDN):", KMDN, "\n")
```

## 2b: Set Noise Threshold

(Lucia slide 12, step 2)

**USER INPUT:** Set the noise threshold multiplier. This is multiplied with the
average noise intensity (KMDN). The default is 0 (removes nothing). You MUST
change this based on your data.

Lucia's example uses 5 (slide 12) or 6-10 (slide 13). Start with 6.

```{r noise-threshold}
## USER INPUT: Assign noise threshold
## This number is multiplied with KMDN to set the absolute threshold.
## Lucia recommends starting at 5-6, check SN plots to verify.
## SET THIS TO A VALUE > 0 or no noise will be removed!
noise_threshold_1 <- 6

## Number of peaks above the threshold
peaks <- Data_4 %>% filter(Abundance > noise_threshold_1 * KMDN)
count(peaks)
```

## 2c: KMD Plot After Noise Removal

(Lucia slide 12, step 3)

Check before and after. The scattered noise "sea" should be gone.

```{r kmd-after}
## New KMD plot with noise removed
peaks_minus_noise <- KMDNoise(peaks)
plot2 <- peaks_minus_noise[["KMD"]]
plot2
```

## 2d: SNplot - Additional Noise Visualization

(Lucia slide 14)

Red peaks = below threshold. Blue peaks = above threshold. Change parameters
to check different mass regions.

**USER INPUT:** Adjust `noise_threshold_2` if you want a different threshold for
the final analysis. Otherwise keep it the same as `noise_threshold_1`.

```{r snplot}
## USER INPUT: Update noise threshold if needed
## Otherwise keep the same as noise_threshold_1
noise_threshold_2 <- noise_threshold_1

## Visualization of noise cutoffs
SNplot(Data, cut = noise_threshold_2 * KMDN, mass = 375, window.x = 500, window.y = 150)
```

**Check different mass regions as needed:**

```{r snplot-extra, eval=FALSE}
## Uncomment and run to check specific regions
# SNplot(Data, cut = noise_threshold_2 * KMDN, mass = 200, window.x = 15, window.y = 10)
# SNplot(Data, cut = noise_threshold_2 * KMDN, mass = 500, window.x = 15, window.y = 10)
```

**USER INPUT:** Change parameter based on how you want to visualize the SN Plot.
Change the parameter until you are satisfied with your data (Lucia slide 12, step 4).

---

# Section 3: Isotope Prescreening (IsoFiltR)

(Lucia slide 16)

Separates monoisotopic peaks from 13C and 34S isotope partners.

- **Carbrat = 60**: 13C peak must be >= ~1.7% of monoisotopic peak to be valid
- **Sulfrat = 30**: 34S intensity ratio threshold
- **Carberr = 5**: mass accuracy for 12C/13C matching (ppm)
- **Sulferr = 5**: mass accuracy for 32S/34S matching (ppm)

These match both the package defaults and Lucia slide 16. Change only if you have reason to.

```{r isofilt}
## Isotope prescreening (Lucia slide 16 - exact parameters)
Isotopes <- IsoFiltR(Data_4,
                     SN = noise_threshold_2 * KMDN,
                     Carbrat = 60,
                     Sulfrat = 30,
                     Sulferr = 5,
                     Carberr = 5)

Mono <- Isotopes[["Mono"]]
Iso <- Isotopes[["Iso"]]
countmono <- count(Mono)
countiso <- count(Iso)
df0 <- data.frame(countmono, countiso) %>%
  rename(Monoisotopes = n, Polyisotopes = n.1)
df0
```

---

# Section 4: Preliminary CHO Assignment (MFAssignCHO)

(Lucia slides 17-22)

MFAssignCHO is a simpler version that only assigns C, H, O formulas.
This is done BEFORE recalibration.

**CRITICAL: Check your ion mode!** Emily's data: is it positive [M+H]+ or
negative [M-H]-?

**Parameter values below are MFAssignR package defaults** (from GitHub source
`skschum/MFAssignR`). Lucia's SOA-specific values are noted where they differ.

If Emily's data is negative mode, change `ionMode` to `"neg"`.

```{r cho-assignment, fig.height=5, fig.width=7, paged.print=FALSE, echo=TRUE}
## CHO assignments
## Parameters: MFAssignR package defaults (verified from GitHub source code)
Assign <- MFAssignCHO_RMD(
  Mono, Iso,
  ionMode = "pos",               # Set for your data's ionization mode
  SN = noise_threshold_2 * KMDN, # Consistent with noise assessment
  ppm_err = 3,                   # Package default: 3 (Lucia used 4)
  NMScut = "on",                 # Package default: "on"
  DeNovo = 1000,                 # Package default: 1000 (Lucia used 750)
  lowMW = 100,                   # Package default: 50; keep 100 for Emily's range
  highMW = 1000,                 # Package default: 1000 (Lucia used 750; Emily has peaks up to 999)
  H_Cmin = 0.3,                  # Package default: 0.3
  H_Cmax = 3,                    # Package default: 3
  O_Cmax = 1.2,                  # Package default: 1.2 (Lucia used 2.5)
  DBEOmax = 13,                  # Package default: 13 (Lucia used 80)
  DBEOmin = -13                  # Package default: -13 (Lucia used -20)
)

## Preliminary assignment plotting
Unambig1.temp <- Assign[["Unambig"]]
Unambig1 <- Unambig1.temp %>% arrange(desc(abundance)) %>% distinct(formula, .keep_all = TRUE)
Ambig1 <- Assign[["Ambig"]]
Unassigned1 <- Assign[["None"]]
MSAssign_CHO <- Assign[["MSAssign"]]
Error_CHO <- Assign[["Error"]]
MSgroups_CHO <- Assign[["MSgroups"]]
VK_CHO <- Assign[["VK"]]

## 4 output plots (Lucia slide 20):
## Mass spectra (red=unassigned, blue=ambiguous, green=assigned)
MSAssign_CHO
## Error plot
Error_CHO
## Assigned mass spectra
MSgroups_CHO
## Van Krevelen Plot
VK_CHO
```

## 4b: Number of Preliminary CHO Assignments

(Lucia slide 22)

```{r cho-counts, results=TRUE, paged.print=TRUE}
countunambig1.temp <- count(Unambig1.temp)
countunambig1 <- count(Unambig1)
countambig1 <- count(Ambig1)
countunassigned1 <- count(Unassigned1)

df <- data.frame(countunambig1.temp, countunambig1, countambig1, countunassigned1) %>%
  rename("Preliminary unambiguous CHO MF" = n,
         "Preliminary unambiguous MF (no duplicates)" = n.1,
         "Preliminary ambiguous CHO MF" = n.2,
         "Preliminary unassigned CHO MF" = n.3)
df
```

---

# Section 5: Mass Recalibration

(Lucia slides 23-27)

## 5a: Recalibrant List

(Lucia slide 24, step 1)

RecalList ranks homologous series from the CHO assignment by their suitability
as recalibrants.

```{r recallist}
## Recalibrant list (Lucia slide 24)
RecalList <- RecalList(df = Unambig1)
```

## 5b: Auto-Select Recalibration Series

Selects the best 3 series from RecalList automatically using the documented
criteria: lowest Peak_Score, Peak_Distance closest to 1, widest mass range,
and coverage of different mass regions.

```{r auto-select-recal}
## Parse mass range into numeric low/high columns
RL <- as.data.frame(RecalList)
RL$mass_low <- as.numeric(sub("-.*", "", RL$`Mass Range`))
RL$mass_high <- as.numeric(sub(".*-", "", RL$`Mass Range`))
RL$mass_span <- RL$mass_high - RL$mass_low

## Score each series: lower is better
## Combine Peak_Score (closer to 0 = better) and Peak_Distance deviation from 1
RL$quality <- RL$`Peak Score` + abs(RL$`Peak Distance` - 1)

## Require at least 4 observed peaks and mass span > 50 Da
candidates <- RL %>%
  filter(`Number Observed` >= 4, mass_span > 50) %>%
  arrange(quality)

## Selection strategy (from MFAssignR docs + Galaxy tutorial):
## Take 6 best-quality series (lowest Peak Score + Peak Distance deviation).
## Recal() divides spectrum into mzRange-wide bins, each needs >2 recalibrants.
## Docs say to increase mzRange up to 80 if gaps occur (default 30 is too narrow).
## We use mzRange=80 in the Recal call below (Galaxy tutorial uses 50).
selected <- head(candidates, 6)

cat("Auto-selected recalibration series:\n")
print(selected[, c("Series", "Number Observed", "Mass Range", "Peak Score",
                    "Peak Distance", "quality")])

series_names <- selected$Series
cat("\nSeries names:", paste(series_names, collapse = ", "), "\n")
cat("Combined mass range:",
    round(min(selected$mass_low), 1), "-",
    round(max(selected$mass_high), 1), "Da\n")
```

## 5c: Run Recalibration

(Lucia slides 25-26)

```{r recal, fig.height=5, fig.width=7}
## Recalibration using auto-selected series
## mzRange=80: MFAssignR docs say "increase to even values as 80" if gap error occurs.
## Galaxy tutorial uses mzRange=50. Default 30 is too narrow for our data.
n_series <- length(series_names)
Recalibration <- Recal(
  Unambig1,
  peaks = Mono,
  isopeaks = Iso,
  mode = "pos",
  mzRange = 80,
  SN = noise_threshold_2 * KMDN,
  series1 = if (n_series >= 1) series_names[1] else NA,
  series2 = if (n_series >= 2) series_names[2] else NA,
  series3 = if (n_series >= 3) series_names[3] else NA,
  series4 = if (n_series >= 4) series_names[4] else NA,
  series5 = if (n_series >= 5) series_names[5] else NA,
  series6 = if (n_series >= 6) series_names[6] else NA,
  series7 = if (n_series >= 7) series_names[7] else NA,
  series8 = if (n_series >= 8) series_names[8] else NA,
  step_O = 3,
  step_H2 = 5,
  CalPeak = 150
)
```

**CHECK (Lucia slide 24):** The output prints two numbers. The error should
DECREASE after recalibration. If it increased, re-examine the selected series.

## 5d: Recalibration Plotting

(Lucia slide 27)

```{r recal-plot, fig.height=5, fig.width=4.5}
## Recalibration plotting (Lucia slide 27)
Plot_Recalibration <- Recalibration[["Plot"]]
Plot_Recalibration

## Extract recalibrated peaks
Mono2 <- Recalibration[["Mono"]]
Iso2 <- Recalibration[["Iso"]]
List <- Recalibration[["RecalList"]]

cat("Recalibrated monoisotopic peaks:", nrow(Mono2), "\n")
cat("Recalibrated isotope peaks:", nrow(Iso2), "\n")
```

**CHECK (Lucia slide 27):** Mass spectrum with recalibration series highlighted
in blue. It's nearly impossible to cover the whole range. That's fine as long
as you see the error decrease.

---

# Section 6: Final MF Assignment with Heteroatoms

(Lucia slides 28-30)

MFAssign does the final assignment using RECALIBRATED masses. Now includes
N, S, P heteroatoms.

**Parameter values below are MFAssignR package defaults** (from GitHub source
`skschum/MFAssignR`). Lucia's SOA-specific values are noted where they differ.

**NOTE on Nx/Sx/Px (heteroatoms):** Package defaults are all 0 (CHO only).
For plant metabolomics with amino acids, glucosinolates, phospholipids, etc.,
you likely need Nx > 0, Sx > 0, and/or Px > 0. D'Andrilli et al. 2013 used
Nx=2 and Sx=2. **Discuss with your PI before changing these.**

```{r final-assignment, fig.height=5, fig.width=7, paged.print=FALSE, echo=TRUE}
## Final assignments - MFAssignR package defaults (verified from GitHub source)
Assign <- MFAssign_RMD(
  peaks = Mono2,
  isopeaks = Iso2,
  lowMW = 100,                   # Package default: 50; keep 100 for Emily's range
  highMW = 1000,                 # Package default: 1000 (Lucia used 750; Emily has peaks up to 999)
  ionMode = "pos",               # Set for your data's ionization mode
  SN = noise_threshold_2 * KMDN,
  ppm_err = 3,                   # Package default: 3
  iso_err = 3,                   # Package default: 3 (Lucia used 1.5)
  H_Cmin = 0.3,                  # Package default: 0.3
  O_Cmax = 1.2,                  # Package default: 1.2 (Lucia used 2.5)
  H_Cmax = 3,                    # Package default: 3
  DBEOmax = 13,                  # Package default: 13 (Lucia used 25)
  DBEOmin = -13,                 # Package default: -13 (Lucia used -20)
  HetCut = "off",                # Package default: "off"
  NMScut = "on",                 # Package default: "on"
  SulfCheck = "on",              # Package default: "on"
  Ambig = "off",                 # Package default: "off"
  MSMS = "off",                  # Package default: "off"
  N3corr = "on",                 # Package default: "on"
  DeNovo = 300,                  # Package default: 300 (Lucia used 350)
  Nx = 5,                        # Prior run: 667 formulas hit N=4 ceiling; raised to 5
  Sx = 2,                        # Prior run: 0 S-containing formulas found; keep at 2 in case
  Px = 3                         # Prior run: 260 formulas hit P=2 ceiling; raised to 3
)

## Extract results
Unambig2 <- Assign[["Unambig"]]
Ambig2 <- Assign[["Ambig"]]
Unassigned2 <- Assign[["None"]]
MSAssign_MF <- Assign[["MSAssign"]]
Error_MF <- Assign[["Error"]]
MSgroups_MF <- Assign[["MSgroups"]]
VK_MF <- Assign[["VK"]]
KMD_MF <- Assign[["KMD"]]
MSgroupstwo_MF <- Assign[["MSgroupstwo"]]
DBEO_MF <- Assign[["DBEO"]]
VK2_MF <- Assign[["VK2"]]

## 4 output plots (same as CHO step - Lucia slide 29)
MSAssign_MF
Error_MF
MSgroups_MF
VK_MF
```

---

# Section 7: Final Unambiguous Assignments

(Lucia slide 31)

Run this chunk (no user input needed).

```{r final-counts, results=TRUE, paged.print=TRUE}
## Final unambiguous assignments (Lucia slide 31)
All.unambig.temp <- Unambig2
All.unambig <- All.unambig.temp %>%
  arrange(desc(abundance)) %>%
  distinct(formula, .keep_all = TRUE)
All.ambig <- Ambig2

countallunambig.temp <- count(All.unambig.temp)
countallunambig <- count(All.unambig)
countallambig <- count(All.ambig)
countallunassigned <- count(Unassigned2)

## Isotope validation (Lucia slide 31)
C13isotopes <- All.unambig %>% filter(C13_mass != 0)
countC13 <- count(C13isotopes)
C13isotopes2 <- All.unambig %>% filter(C13_mass2 != 0)
countC13two <- count(C13isotopes2)
S34isotopes <- All.unambig %>% filter(S34_mass != 0)
countS34 <- count(S34isotopes)

df2 <- data.frame(countallunambig.temp, countallunambig, countallambig,
                   countallunassigned, countC13, countC13two, countS34) %>%
  rename("Monoisotopic MF (duplicates included)" = n,
         "Final monoisotopic MF (duplicates removed)" = n.1,
         "Number of ambiguous MF" = n.2,
         "Masses not assigned an MF" = n.3,
         "13C isotopologues" = n.4,
         "13C2 isotopologues" = n.5,
         "34S isotopologues" = n.6)
df2

Percent_Assigned <- (countallunambig.temp + countC13 + countC13two + countS34) /
  (count(Mono2) + count(Iso2)) * 100
df3 <- data.frame(Percent_Assigned) %>%
  rename("Percent of Peaks Assigned" = n)
df3

## Composition breakdown
All.unambig$group <- factor(All.unambig$group,
                            levels = c("CHO", "CHNO", "CHOS", "CHNOS",
                                       "CH", "CHN", "CHNS", "CHS",
                                       "CHOP", "CHNOP"))
Unambigstat <- All.unambig %>%
  group_by(group) %>%
  summarize(total = n())
Unambigstat
```

---

# Section 8: Bulk Properties

(Lucia slide 33)

```{r bulk-properties, results=TRUE}
average <- All.unambig %>%
  summarise(" " = "Average",
            "O/C" = mean(O_C), "H/C" = mean(H_C), "DBE" = mean(DBE),
            "#C" = mean(C), "#O" = mean(O), "#N" = mean(N),
            "#S" = mean(S), "#P" = mean(P),
            "Absolute error" = mean(AE_ppm), "m/z" = mean(exp_mass))

sd <- All.unambig %>%
  summarise(" " = "Standard deviation",
            "O/C" = sd(O_C), "H/C" = sd(H_C), "DBE" = sd(DBE),
            "#C" = sd(C), "#O" = sd(O), "#N" = sd(N),
            "#S" = sd(S), "#P" = sd(P),
            "Absolute error" = sd(AE_ppm), "m/z" = sd(exp_mass))

table1 <- bind_rows(average, sd)
table1
```

---

# Section 9: Save Output Files

(Lucia slide 32)

```{r save-files}
## Write files (Lucia slide 32 - exact format)
max.abundance <- max(Data$Abundance)

All.unambig <- All.unambig %>%
  mutate(SNR = abundance / KMDN, n_abundance = abundance / max.abundance)
write.csv(All.unambig, "formulas_assigned.csv", row.names = FALSE)

if (nrow(All.ambig) > 0) {
  All.ambig <- All.ambig %>%
    mutate(SNR = abundance / KMDN, n_abundance = abundance / max.abundance)
}
write.csv(All.ambig, "formulas_ambiguous.csv", row.names = FALSE)

Unassigned2 <- Unassigned2 %>%
  mutate(SNR = abundance / KMDN, n_abundance = abundance / max.abundance,
         tag = "Unassigned")
write.csv(Unassigned2, "formulas_unassigned.csv", row.names = FALSE)

all <- bind_rows(All.unambig, All.ambig, Unassigned2)
write.csv(all, "formulas_full.csv", row.names = FALSE)

cat("Saved:\n")
cat("  formulas_assigned.csv    (UMF - unambiguous)\n")
cat("  formulas_ambiguous.csv   (AMF - ambiguous)\n")
cat("  formulas_unassigned.csv  (NoMF - not assigned)\n")
cat("  formulas_full.csv        (Full - all combined)\n")
```

---

# Section 10: z=2 Pass (Doubly Charged Peaks)

24% of Emily's peaks are doubly charged (z=2). MFAssignR defaults to z=1,
which would give wrong neutral masses for these peaks. This section runs the
full pipeline again on z=2 peaks only, then combines results.

**If you have no z=2 peaks, skip this section.**

## 10a: Run MFAssignR Pipeline for z=2

This repeats the core pipeline (KMDNoise → IsoFiltR → MFAssignCHO → Recal →
MFAssign) for doubly charged peaks. You will need to:
1. Re-examine the KMD noise plot (may differ from z=1)
2. Pick new recalibration series from the z=2 RecalList
3. Verify the diagnostic plots

```{r z2-check}
cat("z=2 peaks available:", nrow(Data_z2), "\n")
if (nrow(Data_z2) == 0) {
  cat("No z=2 peaks — skipping Section 10.\n")
}
```

```{r z2-pipeline, eval=(nrow(Data_z2) > 0)}
## ---- KMDNoise for z=2 ----
Noise_z2 <- KMDNoise(Data_z2)
Noise_z2[["KMD"]]
KMDN_z2 <- Noise_z2[["Noise"]]
cat("z=2 KMD Noise level:", KMDN_z2, "\n")

## USER INPUT: Adjust noise threshold for z=2 if needed
noise_threshold_z2 <- noise_threshold_1  # Start with same as z=1

peaks_z2 <- Data_z2 %>% filter(Abundance > noise_threshold_z2 * KMDN_z2)
cat("z=2 peaks above noise:", nrow(peaks_z2), "\n")

## ---- IsoFiltR for z=2 ----
Isotopes_z2 <- IsoFiltR(Data_z2,
                         SN = noise_threshold_z2 * KMDN_z2,
                         Carbrat = 60, Sulfrat = 30,
                         Sulferr = 5, Carberr = 5)
Mono_z2 <- Isotopes_z2[["Mono"]]
Iso_z2 <- Isotopes_z2[["Iso"]]
cat("z=2 monoisotopic:", nrow(Mono_z2), "\n")
cat("z=2 polyisotopic:", nrow(Iso_z2), "\n")

## ---- MFAssignCHO for z=2 (Zx=2) ----
## NOTE: Zx=2 tells MFAssignR these are doubly charged peaks
Assign_z2_cho <- MFAssignCHO_RMD(
  Mono_z2, Iso_z2,
  ionMode = "pos",
  SN = noise_threshold_z2 * KMDN_z2,
  ppm_err = 3, NMScut = "on",
  DeNovo = 1000, lowMW = 100, highMW = 1000,
  H_Cmin = 0.3, H_Cmax = 3, O_Cmax = 1.2,
  DBEOmax = 13, DBEOmin = -13,
  Zx = 2                         # DOUBLY CHARGED
)

Unambig1_z2_temp <- Assign_z2_cho[["Unambig"]]
Unambig1_z2 <- Unambig1_z2_temp %>% arrange(desc(abundance)) %>% distinct(formula, .keep_all = TRUE)
cat("z=2 preliminary CHO assignments:", nrow(Unambig1_z2), "\n")

## Diagnostic plots for z=2 CHO
Assign_z2_cho[["MSAssign"]]
Assign_z2_cho[["Error"]]
Assign_z2_cho[["VK"]]
```

## 10b: z=2 Recalibration (Auto-Selected)

```{r z2-recal, eval=(nrow(Data_z2) > 0)}
## RecalList for z=2
RecalList_z2 <- RecalList(df = Unambig1_z2)

## Auto-select series using same logic as z=1
RL_z2 <- as.data.frame(RecalList_z2)
RL_z2$mass_low <- as.numeric(sub("-.*", "", RL_z2$`Mass Range`))
RL_z2$mass_high <- as.numeric(sub(".*-", "", RL_z2$`Mass Range`))
RL_z2$mass_span <- RL_z2$mass_high - RL_z2$mass_low
RL_z2$quality <- RL_z2$`Peak Score` + abs(RL_z2$`Peak Distance` - 1)

candidates_z2 <- RL_z2 %>%
  filter(`Number Observed` >= 4, mass_span > 50) %>%
  arrange(quality)

selected_z2 <- head(candidates_z2, 6)
cat("z=2 auto-selected series:\n")
print(selected_z2[, c("Series", "Number Observed", "Mass Range", "Peak Score",
                       "Peak Distance", "quality")])
series_names_z2 <- selected_z2$Series
n_z2 <- length(series_names_z2)

Recalibration_z2 <- Recal(
  Unambig1_z2,
  peaks = Mono_z2,
  isopeaks = Iso_z2,
  mode = "pos",
  mzRange = 80, CalPeak = 150,
  SN = noise_threshold_z2 * KMDN_z2,
  series1 = if (n_z2 >= 1) series_names_z2[1] else NA,
  series2 = if (n_z2 >= 2) series_names_z2[2] else NA,
  series3 = if (n_z2 >= 3) series_names_z2[3] else NA,
  series4 = if (n_z2 >= 4) series_names_z2[4] else NA,
  series5 = if (n_z2 >= 5) series_names_z2[5] else NA,
  series6 = if (n_z2 >= 6) series_names_z2[6] else NA,
  step_O = 3, step_H2 = 5
)

Recalibration_z2[["Plot"]]
Mono2_z2 <- Recalibration_z2[["Mono"]]
Iso2_z2 <- Recalibration_z2[["Iso"]]
cat("z=2 recalibrated monoisotopic:", nrow(Mono2_z2), "\n")
```

## 10c: z=2 Final Assignment

```{r z2-final, eval=(nrow(Data_z2) > 0)}
## MFAssignR KNOWN ISSUES: "Only singly charged compounds can be assigned."
## MFAssign/MFAssign_RMD crash with Zx=2 (object 'E' not found).
## MFAssignCHO_RMD handles Zx=2 correctly, so we re-run CHO on recalibrated peaks.
## This gives CHO-only formulas for z=2 (no N/S/P), which is the best we can do.
Assign_z2_final <- MFAssignCHO_RMD(
  Mono2_z2, Iso2_z2,
  ionMode = "pos",
  SN = noise_threshold_z2 * KMDN_z2,
  ppm_err = 3, NMScut = "on",
  DeNovo = 1000, lowMW = 100, highMW = 1000,
  H_Cmin = 0.3, H_Cmax = 3, O_Cmax = 1.2,
  DBEOmax = 13, DBEOmin = -13,
  Zx = 2
)

Unambig2_z2 <- Assign_z2_final[["Unambig"]] %>%
  arrange(desc(abundance)) %>%
  distinct(formula, .keep_all = TRUE)

cat("z=2 CHO assignments (recalibrated):", nrow(Unambig2_z2), "\n")

Assign_z2_final[["MSAssign"]]
Assign_z2_final[["Error"]]
Assign_z2_final[["VK"]]
```

## 10d: Combine z=1 and z=2 Results

```{r combine-results}
## Add charge state column to track which pass each formula came from
All.unambig$charge <- 1

if (nrow(Data_z2) > 0 && exists("Unambig2_z2") && nrow(Unambig2_z2) > 0) {
  ## Process z=2 results (same as Section 7)
  All.unambig.z2 <- Unambig2_z2 %>%
    arrange(desc(abundance)) %>%
    distinct(formula, .keep_all = TRUE)
  All.unambig.z2$charge <- 2

  max.abundance.z2 <- max(Data_z2$Abundance)
  All.unambig.z2 <- All.unambig.z2 %>%
    mutate(SNR = abundance / KMDN_z2, n_abundance = abundance / max.abundance.z2)

  ## Combine z=1 and z=2
  All.unambig.combined <- bind_rows(All.unambig, All.unambig.z2)

  cat("\n=== COMBINED RESULTS ===\n")
  cat("z=1 formulas:", nrow(All.unambig), "\n")
  cat("z=2 formulas:", nrow(All.unambig.z2), "\n")
  cat("Total:", nrow(All.unambig.combined), "\n")
} else {
  All.unambig.combined <- All.unambig
  cat("No z=2 results to combine. Using z=1 only.\n")
}

## Overwrite the output CSV with combined results
write.csv(All.unambig.combined, "formulas_assigned.csv", row.names = FALSE)
cat("\nSaved combined results to formulas_assigned.csv\n")
```

---

# NEXT STEPS

After completing this notebook:

1. Run `merge_formulas.R` to merge assignments back to the original data
2. This produces `Emily_Data_WITH_FORMULAS.xlsx` for the dashboard
3. Run the dashboard pipeline
